---
title: "FARS Crash Fatalities Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    source_code: "https://github.com/jshtulman1/fars-dashboard"
    theme: cosmo
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(shiny)

# Load data
fars <- read_csv("fars_data.csv")

# Convert to long format
fars_long <- fars %>%
  pivot_longer(
    cols = -c(state_abbr, state_name, population_2023),
    names_to = c(".value", "year"),
    names_pattern = "(.*)_(\\d{4})"
  ) %>%
  mutate(
    year = as.integer(year),
    rate_per100k = round(fatalities / population_2023 * 100000, 1)
  )

# Create vector containing state name choices
state_choices <- setNames(fars$state_abbr, fars$state_name)
```

# Overview

Row
-----------------------------------
### Total Traffic Fatalities (2023)
```{r}
total_fat <- sum(fars$fatalities_2023)
total_fat_2022 <- sum(fars$fatalities_2022)
delta_pct <- round((total_fat - total_fat_2022) / total_fat_2022 * 100, 1)

valueBox(
  value = format(total_fat, big.mark = ","),
  caption = paste0("<b>Total Traffic Fatalities</b><br>Down ", abs(delta_pct), "% from 2022"),
  icon = "fa-car-crash",
  color = "danger"
)
```

### Fatal Crashes (2023)
```{r}
# NHTSA published 37,654 fatal crashes in 2023, 39,221 in 2022
fatal_crashes_2023 <- 37654
fatal_crashes_2022 <- 39221
crashes_delta <- round((fatal_crashes_2023 - fatal_crashes_2022) / fatal_crashes_2022 * 100, 1)

valueBox(
  value = format(fatal_crashes_2023, big.mark = ","),
  caption = paste0("<b>Fatal Crashes</b><br>Down ", abs(crashes_delta), "% from 2022"),
  icon = "fa-exclamation-triangle",
  color = "warning"
)
```

### National Rate per 100,000 People
```{r}
natl_rate <- round(total_fat / sum(fars$population_2023) * 100000, 1)

valueBox(
  value = natl_rate,
  caption = "<b>National Fatality Rate</b><br>per 100,000 residents",
  icon = "fa-users",
  color = "primary"
)
```

### Highest State Rate
```{r}
top_state <- fars_long %>%
  filter(year == 2023) %>%
  slice_max(rate_per100k, n = 1)

valueBox(
  value = top_state$rate_per100k,
  caption = paste0("<b>Highest State Rate</b><br>", top_state$state_name, " (per 100k)"),
  icon = "fa-arrow-up",
  color = "danger"
)
```

### Lowest State Rate
```{r}
bottom_state <- fars_long %>%
  filter(year == 2023) %>%
  slice_min(rate_per100k, n = 1)

valueBox(
  value = bottom_state$rate_per100k,
  caption = paste0("<b>Lowest State Rate</b><br>", bottom_state$state_name, " (per 100k)"),
  icon = "fa-arrow-down",
  color = "success"
)
```

Row {data-height=650}
------------------
### Choropleth Map

```{r}
# Map Inpputs
fillCol(flex = c(NA, 1),
  fillRow(flex = c(1, 1),
    selectInput("map_year", "Year:",
                choices = c("2023", "2022"),
                selected = "2023"),
    selectInput("map_metric", "Metric:",
                choices = c("Total Fatalities" = "fatalities",
                            "Rate per 100,000 People" = "rate_per100k"),
                selected = "fatalities")
  ),
  plotlyOutput("map_plot")
)

output$map_plot <- renderPlotly({
  
  # Filter the data by year
  map_data <- fars_long %>%
    filter(year == as.integer(input$map_year))
  
  # Plot selected metric
  if (input$map_metric == "fatalities") {
    map_data$z_value <- map_data$fatalities
    legend_title <- "Fatalities"
  } else {
    map_data$z_value <- map_data$rate_per100k
    legend_title <- "Per 100k"
  }
  
  # Hover box
  map_data <- map_data %>%
    mutate(
      hover_text = paste0(
        state_name, "<br>",
        "Fatalities: ", format(fatalities, big.mark = ","), "<br>",
        "Rate per 100k: ", rate_per100k
      )
    )
  
  # Build Chloropleth
  plot_geo(map_data, locationmode = "USA-states") %>%
    add_trace(
      z = ~z_value,
      locations = ~state_abbr,
      text = ~hover_text,
      hoverinfo = "text",
      color = ~z_value,
      colors = "Greens"
    ) %>%
    colorbar(title = legend_title) %>%
    layout(
      geo = list(scope = "usa"),
      margin = list(t = 75, b = 5, l = 5, r = 5)
    ) %>%
    config(displayModeBar = FALSE)
})
```

### Contributing Factors: State Comparison

```{r}
# Bar Chart Inputs
fillCol(flex = c(NA, 1),
  fillRow(flex = c(1, 1, 1),
    selectInput("bar_year", "Year:",
                choices = c("2023", "2022"),
                selected = "2023"),
    selectInput("state_a", "State A:",
                choices = state_choices,
                selected = "WV"),
    selectInput("state_b", "State B:",
                choices = state_choices,
                selected = "GA")
  ),
  plotlyOutput("bar_plot")
)

output$bar_plot <- renderPlotly({
  
  # Get state data for specified year
  data_a <- fars_long %>%
    filter(state_abbr == input$state_a, year == as.integer(input$bar_year))
  
  data_b <- fars_long %>%
    filter(state_abbr == input$state_b, year == as.integer(input$bar_year))
  
  # Build plot and sort by largest disparity
  factors_df <- tibble(
    factor = c("Alcohol-Impaired", "Speeding-Related", "Unrestrained Occupant"),
    state_a_pct = c(data_a$alcohol_pct, data_a$speeding_pct, data_a$unrestrained_pct),
    state_b_pct = c(data_b$alcohol_pct, data_b$speeding_pct, data_b$unrestrained_pct)
  ) %>%
    mutate(
      disparity = abs(state_a_pct - state_b_pct),
      diff_text = paste0("Difference: ", round(state_a_pct - state_b_pct, 1), " pts")
    ) %>%
    arrange(disparity) %>%
    mutate(factor = factor(factor, levels = factor))
  
  # Legend
  name_a <- data_a$state_name
  name_b <- data_b$state_name
  
  # Bar chart
  plot_ly(factors_df) %>%
    add_trace(
      type = "bar",
      orientation = "h",
      y = ~factor,
      x = ~state_a_pct,
      name = name_a,
      marker = list(color = "rgb(37, 99, 235)"),
      text = ~paste0(state_a_pct, "%"),
      textposition = "outside",
      customdata = ~diff_text,
      hovertemplate = paste0(name_a, ": %{x}%<br>%{customdata}<extra></extra>")
    ) %>%
    add_trace(
      type = "bar",
      orientation = "h",
      y = ~factor,
      x = ~state_b_pct,
      name = name_b,
      marker = list(color = "rgb(234, 88, 12)"),
      text = ~paste0(state_b_pct, "%"),
      textposition = "outside",
      customdata = ~diff_text,
      hovertemplate = paste0(name_b, ": %{x}%<br>%{customdata}<extra></extra>")
    ) %>%
    layout(
      barmode = "group",
      xaxis = list(title = "Percent of Fatal Crashes (%)",
                   ticksuffix = "%",
                   range = list(0, 50)),
      yaxis = list(title = ""),
      legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.25),
      margin = list(t = 75, b = 100)
    ) %>%
    config(displayModeBar = FALSE)
})
```

Row {data-height=120}
-----------------------------------------------------------------------

### Why this matters

Traffic crashes kill more than 40,000 Americans every year, and reducing that toll requires understanding where and why fatalities occur. This dashboard enables researchers and policy analysts to quickly observe trends and make comparisons about traffic fatality patterns between states, supporting the targeted interventions needed to reduce traffic fatalities in the United States.

About the Data
=======================================================================

Row
-----------------------------------------------------------------------

###

**Data Source**

The Fatality Analysis Reporting System (FARS) is a nationwide census of fatal motor vehicle traffic crashes maintained by the National Highway Traffic Safety Administration (NHTSA) since 1975, with aggregate data for this dashboard sourced from NHTSA's State Traffic Data fact sheets for 2023 (https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813743)
and 2022 (https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813592.pdf). 

**Sample Size**

The dashboard covers 50 states and Washington DC, totaling 40,901 traffic fatalities in 2023 and 42,514 in 2022.

**Data Collection**

NHTSA has agreements with a traffic agency in each state where trained analysts provide information on every fatal crash using accident reports, state vehicle registration and driver licensing files, hospital records, death certificates, and coroner reports.

**Study Population**

To qualify as a FARS case, a crash must involve a motor vehicle traveling on a publicly accessible trafficway, and result in the death of a vehicle occupant or non-occupant within 30 days of the crash.

**Time Period**

This dashboard displays data for motor vehicle crashes which was collected in 2022 and 2023. 
